\documentclass[../main.tex]{subfiles}
\begin{document}

\begin{center}
    \Large\textbf{ACKNOWLEDGMENT}
\end{center}
\vspace{1cm}

I would like to express my sincere gratitude to everyone who has supported me throughout the completion of this graduation project.

First and foremost, I would like to thank my supervisor, \textbf{MSc. Bui Trong Tung}, for his dedicated guidance, valuable advice, and continuous encouragement. His patience and support helped me overcome many difficulties during this project. I am especially grateful for his kindness and understanding of my work commitments. Also, I would like to thank the BKCS Center for providing the facilities and resources that made this project possible.

My sincere thanks go to Hanoi University of Science and Technology and all of my lecturers. Over the past four years, they have shared their knowledge, experience, and valuable lessons, which have helped me grow both academically and personally.

Finally, I would like to express my deepest gratitude to my family, friends, and colleagues. Their love, encouragement, understanding, and constant support have always motivated me to keep moving forward, especially during challenging times.

Thank you all for being an important part of this journey.

Despite my best efforts, this thesis may still contain some shortcomings due to time constraints and my limited research experience. I sincerely welcome comments and constructive suggestions from professors, colleagues, and friends so that I can continue to improve my work.


\end{document}
