\documentclass[../main.tex]{subfiles}
\begin{document}

\section{Evaluation objectives and rule-set scope}
\label{sec:eval-objectives}

This chapter reports quantitative and qualitative evaluation of the Suricata rule set (\texttt{s7comm.rules}, \texttt{ics\_dos.rules}, \texttt{s7comm\_malformed.rules}) on the lab described in Section~\ref{sec:lab-setup}. Four questions guide the experiments:

\begin{enumerate}
    \item \textbf{Control-path impact.} The IDS operates in passive mirror mode: it neither drops nor modifies packets. Under this deployment model, Suricata has no direct effect on PLC logic execution. Any operational consequence would come only from human response to alerts, which this thesis does not model.

    \item \textbf{Baseline alert behaviour.} On traffic that contains no attack scripts, how often do rules fire, and which SIDs dominate? This thesis reports \emph{alert density} and \emph{operational false positives} rather than a classical false-positive rate, because benign S7comm sessions (Snap7 block listing) are not fully labelled at the packet level.

    \item \textbf{Detection coverage.} For each scripted attack scenario, does at least one expected SID appear in \texttt{fast.log} after mirrored traffic reaches the sensor?

    \item \textbf{Sensor capacity.} Under baseline and DoS load, does Suricata keep up with the mirrored packet stream? The primary kernel-level indicator is \texttt{capture.kernel\_drops}; process CPU and RAM provide secondary evidence.
\end{enumerate}

\textbf{Scope.} The testbed uses one OpenPLC instance, one Attacker VM, and one IDS VM. It does not reproduce continuous HMI polling over days or weeks. Reported figures are reproducible lab evidence (Section~\ref{sec:eval-discussion}), not universal bounds for every OT network.

\section{Evaluation setup on the lab environment}
\label{sec:eval-setup}

\subsection{Topology and role of each virtual machine}

S7comm between the Attacker and the PLC (\texttt{192.168.50.10}:102) traverses the Controller LAN segment (\texttt{192.168.50.0/24}). A mirror copy of that traffic is delivered to \texttt{ens33} on the IDS (\texttt{172.16.16.7}). Suricata inspects the copy only; the Attacker (\texttt{172.16.16.5}) executes scenarios over SSH. Suricata \texttt{HOME\_NET} includes \texttt{192.168.50.0/24} and \texttt{192.168.60.0/24}; the NAT management subnet \texttt{172.16.16.0/24} is excluded.

Table~\ref{tab:eval-lab-setup} lists fixed lab parameters. Tables~\ref{tab:eval-baseline}--\ref{tab:eval-performance} are regenerated by \texttt{detect/eval\_rules.py} whenever rules or scenarios change.

\IfFileExists{../../detect/eval_results/eval_tables_en.tex}{%
  \input{../../detect/eval_results/eval_tables_en.tex}%
}{%
\begin{table}[H]
    \centering
    \footnotesize
    \begin{tabular}{|l|l|}
        \hline
        \textbf{Component} & \textbf{Lab value} \\
        \hline
        IDS (Suricata) & 172.16.16.7 (lubuntu), mirror ens33 \\
        Attacker & 172.16.16.5 (ubuntu), repo /home/ubuntu/Thesis \\
        PLC (OpenPLC :102) & 192.168.50.10 (rack 0, slot 2) \\
        HOME\_NET & 192.168.50.0/24, 192.168.60.0/24 \\
        \hline
    \end{tabular}
    \caption{Evaluation environment setup (run \texttt{python detect/eval\_rules.py} to refresh dynamic tables)}
    \label{tab:eval-lab-setup}
\end{table}
}

\subsection{Three layers of evaluation criteria}

Evaluation is organised in three complementary layers: scenario detection, baseline alert behaviour, and capture performance.

\paragraph{Layer 1 --- Scenario detection rate.}
For each entry in \texttt{attacks/}, the script checks whether \texttt{fast.log} contains at least one alert whose SID belongs to the expected set (Table~\ref{tab:attack-eval-summary}). A scenario \emph{passes} when \texttt{detected = true}, i.e.\ the intersection of observed and expected SIDs is non-empty. The aggregate metric is
\begin{equation}
    R_{\mathrm{det}} = \frac{\#\{\text{scenarios with detected=true}\}}{N_{\mathrm{scenarios}}}.
\end{equation}
This measures rule \emph{coverage} of scripted attacks, not precision against unlabelled background traffic.

\paragraph{Layer 2 --- Baseline alert density and operational false positives.}
During interval $T$ (default 90\,s), only \texttt{scanBlock.py --list-only} runs on a fixed schedule---valid Snap7 reads without nmap, writes, DoS, or malformed scripts. Every alert that appears in this window is counted as an \emph{operational alert}. Alerts that reflect legitimate protocol use (e.g.\ Setup Communication when Snap7 opens a session) are classified as \emph{operational false positives} in the SOC sense: the signature matched correctly, but no attack was intended.

The reported rate is alert density, not a confusion-matrix false-positive rate:
\begin{equation}
    \label{eq:baseline-density}
    \rho_{\mathrm{baseline}} = \frac{N_{\mathrm{alert\,baseline}}}{T} \times 3600 \quad \mathrm{(alerts/hour)}.
\end{equation}
Because $N_{\mathrm{alert\,baseline}}$ counts rule firings rather than misclassified flows, $\rho_{\mathrm{baseline}}$ must be interpreted together with SID labels (Table~\ref{tab:eval-baseline-sid}).

\paragraph{Layer 3 --- Capture and processing performance.}
From \texttt{stats.log} and \texttt{suricata - top}, the thesis records \texttt{capture.kernel\_packets}, \texttt{capture.kernel\_drops}, and Suricata CPU/RAM. \texttt{kernel\_drops} counts packets the Linux capture path discarded before Suricata's userspace engine could decode them---a direct indicator of sensor overload. At lab load, \texttt{kernel\_drops = 0} means the IDS kept pace with the mirror feed; non-zero drops under DoS would suggest capacity limits (Chapter~6).

\subsection{Automated data collection}

The pipeline is implemented in \texttt{detect/eval\_rules.py}:
\begin{enumerate}
    \item SSH connectivity checks on the IDS and Attacker; verify TCP port 102 on the PLC from the Attacker.
    \item \textbf{Baseline window} of length $T$ with periodic Snap7 reads; count alerts whose timestamps fall inside the window.
    \item \textbf{Attack sequence} (fixed order): recon nmap, recon blocks, write DB, DoS all, malformed all, PLC stop, PLC start---the same commands as the Attack Tool GUI (\texttt{.venv/bin/python} on the Attacker). Expected SIDs can be cross-checked in the Rules Tool alert panel.
    \item Export \texttt{eval\_report.json}, \texttt{eval\_detection.csv}, \texttt{eval\_tables.tex}, \texttt{eval\_report.md} under \texttt{detect/eval\_results/}.
\end{enumerate}

Commands:
\begin{packetdump}
python detect/deploy_ids.py          # once, after rule changes
python detect/eval_rules.py          # baseline 90s + 7 scenarios
python detect/eval_rules.py --baseline-sec 120 --skip-baseline  # attacks only
\end{packetdump}

Single-scenario smoke test: \texttt{python detect/run\_attack\_tests.py -s dos\_all}.

\section{Impact on normal operation}
\label{sec:eval-normal-impact}

\paragraph{No inline effect on the PLC.}
Suricata never sits in the control path. OpenPLC continues to accept S7comm from the Attacker and HMI whether or not alerts are generated. When DoS experiments slow the PLC, the cause is attack traffic volume, not IDS rule logic.

\paragraph{Signatures that also match legitimate engineering traffic.}
Rules \texttt{1000010}, \texttt{1000011} (Write Var) and \texttt{1000030}--\texttt{1000033} (Start/Stop) will fire on authorised maintenance if the engineering station is not whitelisted. In the read-only baseline, alerts were dominated by \texttt{1000001} (Setup) and occasionally \texttt{1000005} (List Blocks). Static signatures cannot distinguish intent without asset context; mitigation requires IP allow lists, change windows, or correlation with maintenance tickets.

\paragraph{Rate-based DoS rules.}
Rules \texttt{1000040}--\texttt{1000043} use Suricata \texttt{threshold} over a 10\,s sliding window (Table~\ref{tab:eval-dos-thresholds}). Occasional Setup or Read SZL traffic stays below these counts; sustained floods exceed them. Consequently, DoS signatures did not fire during the baseline run---an expected outcome when thresholds exceed normal poll rates.

\section{Baseline and operational false-positive results}
\label{sec:eval-fp}

This section reports Layer~2 metrics from Section~\ref{sec:eval-setup}. The latest automated run used a 60\,s baseline window and two \texttt{scanBlock --list-only} invocations:
\begin{itemize}
    \item \textbf{3} alerts recorded; scaled density $\rho_{\mathrm{baseline}} = 180$ alerts/hour (Equation~\eqref{eq:baseline-density}).
    \item All three alerts carried SID \texttt{1000001} (Setup Communication). Snap7 opens an S7 session before listing blocks; the signature matched correctly. These events are \emph{operational false positives} relative to an attack label, not detection errors.
    \item No \texttt{100004x} alerts appeared during baseline, confirming that DoS \texttt{threshold} counts sit above benign read frequency in this configuration.
\end{itemize}

The next section turns to Layer~1: scripted attack scenarios and $R_{\mathrm{det}}$.

\section{Attack detection matrix}
\label{sec:eval-detection}

Layer~1 evaluation asks whether each scripted scenario produces at least one expected SID. Across seven automated scenarios, \textbf{$R_{\mathrm{det}} = 7/7 = 100\%$}: each run yielded at least one expected alert in \texttt{fast.log}. The DoS script generated thousands of alerts (\texttt{1000040}, \texttt{1000042}, among others). Malformed traffic triggered \texttt{1000053} and \texttt{1000054}; not all nine malformed phases mapped to distinct SIDs on OpenPLC because of stack timeouts and partial parsing.

Table~\ref{tab:eval-performance} (Layer~3) shows \texttt{capture.kernel\_packets} rising from $\approx 1{,}13 \times 10^5$ at baseline end to $\approx 1{,}58 \times 10^5$ after the attack batch, with \texttt{kernel\_drops} remaining zero---Suricata processed the full mirrored load in this lab.

Scenario-to-rule mapping appears in Table~\ref{tab:attack-eval-summary}; per-phase DoS mapping in Table~\ref{tab:dos-eval}.

\begin{table}[H]
    \centering
    \footnotesize
    \begin{tabular}{|l|l|l|}
        \hline
        \textbf{Attack group} & \textbf{Script / lab tool} & \textbf{Rules (expected SIDs)} \\
        \hline
        S7 reconnaissance & \texttt{nmap s7-info}, \texttt{scanBlock.py} & 1000001--1000005 \\
        \hline
        Process variable write (DB) & \texttt{test\_readwrite\_db\_via\_snap7.py} & 1000010, 1000011 \\
        \hline
        S7comm DoS & \texttt{s7comm\_dos.py --mode all} & 1000040--1000043 \\
        \hline
        Malformed S7comm & \texttt{s7comm\_malformed.py --mode all} & 1000050--1000058 \\
        \hline
        Start/Stop replay & \texttt{start\_stop\_plc.py} & 1000030--1000033 \\
        \hline
        MITM (secondary observation) & \texttt{s7\_intercept.py} + ARP & 1000013 (+ L2) \\
        \hline
        Up/Download & PCAP pending & 1000020--1000026 (rules ready) \\
        \hline
    \end{tabular}
    \caption{Summary of lab scenarios and Suricata rules}
    \label{tab:attack-eval-summary}
\end{table}

\section{DoS detection results on S7comm}
\label{sec:eval-dos}

The DoS scenario exercises Layer~1 coverage for rate-based rules (\texttt{sid:1000040}--\texttt{1000043}). \texttt{s7comm\_dos.py --mode all} executes three 10\,s phases---\texttt{setup\_flood}, \texttt{tcp\_connect}, \texttt{szl\_flood}---separated by 8\,s pauses. Flood detection relies on \texttt{threshold} counters rather than opcode content alone. Opcode rules such as \texttt{1000001} and \texttt{1000002} may co-fire because the flooded PDUs remain syntactically valid; they supplement but do not replace rate-based flood alerts.

\begin{table}[H]
    \centering
    \footnotesize
    \begin{tabular}{|c|l|l|l|}
        \hline
        \textbf{Step} & \textbf{Phase} & \textbf{Behavior} & \textbf{Suricata SID} \\
        \hline
        1 & \texttt{setup\_flood} & Repeated Setup PDU on new TCP sessions & 1000041, 1000043 \\
        \hline
        2 & \texttt{tcp\_connect} & Rapid connect/disconnect on :102 & 1000040 \\
        \hline
        3 & \texttt{szl\_flood} & \texttt{read\_szl(0x0011)} & 1000042 \\
        \hline
    \end{tabular}
    \caption{Mapping of lab DoS scenario to detection rules}
    \label{tab:dos-eval}
\end{table}

\section{Malformed packet detection results}
\label{sec:eval-malformed}

\texttt{s7comm\_malformed.py --mode all} sends nine structurally invalid variants (TPKT/COTP, S7 header, function layer, truncation) with \texttt{--repeats 2} and \texttt{--pause 6}s between phases. A phase \emph{passes} when $\geq 1$ alert with SID \texttt{100005x} appears. Rule \texttt{1000050} matches any non-\texttt{03~00} TPKT version on port~102; if \texttt{HOME\_NET} is too broad, unrelated traffic may trigger it---production deployments should narrow the monitored subnet.

\section{Suricata performance on the IDS}
\label{sec:eval-performance}

Layer~3 metrics assess whether the sensor keeps pace with the mirror feed. Performance snapshots were taken at baseline start/end, immediately before the attack batch, during DoS, and after all scenarios (Table~\ref{tab:eval-performance} in \texttt{eval\_tables\_en.tex}). Interpretation criteria for this lab:
\begin{itemize}
    \item \texttt{capture.kernel\_drops = 0} under baseline and post-attack load: no kernel-level packet loss on the mirror interface.
    \item Elevated CPU during DoS with continued \texttt{100004x} alerts: threshold rules remain effective under moderate stress.
    \item Maximum sustainable throughput (Mbps or pps) was not measured; capacity planning for plant-scale traffic is out of scope.
\end{itemize}

\section{Discussion and limitations}
\label{sec:eval-discussion}

\begin{itemize}
    \item \textbf{Visibility.} False negatives occur if attack traffic never crosses the mirror path, regardless of signature quality.
    \item \textbf{Baseline duration.} A 60--90\,s read-only window underestimates alert rates under continuous HMI polling; $\rho_{\mathrm{baseline}}$ is a short-window estimate.
    \item \textbf{Intent-blind signatures.} Write, Start, and Stop rules require procedural triage; automatic blocking is inappropriate on OT segments.
    \item \textbf{Upload/download rules.} SIDs \texttt{1000020}--\texttt{1000026} were not validated against lab PCAP because OpenPLC loads programs via REST, not S7 block transfer.
    \item \textbf{Reproducibility.} Re-running \texttt{eval\_rules.py} after \texttt{deploy\_ids.py} regenerates all tables and JSON artefacts in \texttt{detect/eval\_results/}.
\end{itemize}

\end{document}
