\documentclass[../main.tex]{subfiles}
\begin{document}

\section{Summary of the thesis work}

This thesis addressed passive detection of S7comm attack behaviour in a controlled industrial-network laboratory~\cite{nist_sp800_82,veit2021ics_dissectors}. The research question was whether a signature-based network IDS, deployed without touching live control paths, can reliably surface representative S7comm abuses that appear in documented ICS incidents and in the project's own attack simulations. The work proceeded in four linked stages: building a virtual testbed, simulating representative attack scenarios, analysing traffic to derive detection logic, and evaluating a Suricata rule set~\cite{suricata_oisf} against live and replayed traffic. Each stage fed the next: captures from the testbed supplied byte-level evidence for rule design, and the rule set was then exercised on the same traffic classes to verify alert alignment.

\paragraph{Testbed and attack simulation.}
The laboratory replicated a simplified OT segment with an OpenPLC runtime~\cite{autonomylogic_openplc,openplc_s7_driver}, an HMI on a separate subnet, an attacker workstation, and an IDS receiving a mirror copy of Controller-layer traffic. Routing between the Controller and Supervisory subnets allowed the HMI to poll the PLC while the attacker could reach the same S7 endpoint from a different segment---a layout that mirrors common Purdue Level~1/Level~2 separation without reproducing an entire plant. On this testbed, the thesis implemented and documented attack classes that map to common ICS threat patterns~\cite{mitre_ics_attack}:

\begin{itemize}
    \item \textbf{Reconnaissance} over TCP/102, including Nmap \texttt{s7-info} scans, Read SZL enumeration, and Snap7 block inventory;
    \item \textbf{Unauthorized writes} to process datablocks, including the Stuxnet-style change to a setpoint in \texttt{DB1};
    \item \textbf{Denial of service} on the S7comm channel through valid-opcode floods and Setup spam;
    \item \textbf{Malformed PDUs} targeting TPKT, COTP, S7 header, function, and truncation faults;
    \item \textbf{Replay} of Start/Stop control commands captured from engineering tools;
    \item \textbf{Block upload and download} sequences derived from reference Siemens captures;
    \item A \textbf{Stuxnet-inspired man-in-the-middle} scenario~\cite{falliere2010stuxnet,langner2011stuxnet,stuxnet_work_details} that combines an unauthorized write with response interception toward the HMI.
\end{itemize}

Each scenario was scripted, captured with Wireshark, and described at the PDU level in Chapter~3 so that detection logic could be traced back to observable fields rather than to anecdotal attack descriptions.

\paragraph{Rule design and evaluation.}
For each scenario, traffic was analysed to identify stable byte patterns at the TPKT, COTP, S7 header, and function layers. Those patterns were expressed as Suricata signatures with explicit traffic-direction constraints (\texttt{flow:to\_server} vs.\ \texttt{flow:from\_server}) and, where appropriate, rate thresholds for flood detection. The resulting rule set groups related behaviour by SID ranges: session setup and reconnaissance (\texttt{100000x}), read/write variables (\texttt{100001x}), block transfer (\texttt{100002x}), PLC control (\texttt{100003x}), DoS (\texttt{100004x}), and malformed traffic (\texttt{100005x}). Chapter~4 documents the translation principles---anchor bytes, offset~17 vs.\ offset~19 for Job vs.\ Ack\_Data PDUs, and single-match vs.\ threshold rules---that apply uniformly across scenarios.

Evaluation in Chapter~5 applied a three-layer methodology: baseline alert density under benign engineering reads, scenario-level detection coverage against expected rule groups, and sensor processing load under baseline and attack traffic. The campaign exercised ten attack classes---seven live on OpenPLC, two via PCAP replay for block transfer, one for MITM---and confirmed that the rule set reacted to each class while the passive sensor reported no kernel packet drops at the loads generated in the lab. Two desktop tools were added to streamline repeated runs (attack launcher and Suricata rule viewer); they did not change detection logic but made scenario replay and alert inspection reproducible during rule tuning.

\section{Main findings}

The experiments support several conclusions at the level of this thesis scope. The following paragraphs summarise what the lab campaign demonstrated and what it did not claim.

\paragraph{Feasibility of passive S7comm detection.}
Passive, signature-based IDS deployment on a mirrored S7comm segment is feasible for the attack classes studied. Rules anchored on protocol structure---TPKT and COTP headers, \texttt{ROSCTR} values, function codes at offsets~17 and~19---produced alerts that aligned with the intended scenario in every test class. This outcome matters because S7comm lacks built-in authentication: the same byte patterns that appear during legitimate engineering also appear during abuse, so detection must rely on structural and contextual cues rather than on credential failure. The thesis shows that those cues can be encoded explicitly in Suricata rules and verified offline against captured traffic before any inline deployment is considered.

\paragraph{Rate rules and DoS separation.}
Rate-based rules separated DoS floods from isolated legitimate Setup or Read SZL operations when thresholds were set above the baseline poll rate. Setup-only floods triggered \texttt{sid:1000041}; generic Job-header floods triggered \texttt{sid:1000043}; Read SZL abuse triggered \texttt{sid:1000042}. During the 60-second baseline window, none of the DoS rules fired while only Setup alerts from normal Snap7 sessions appeared. The finding indicates that threshold tuning is a practical lever: operators can raise flood counts above expected engineering poll rates without removing opcode-based signatures that remain useful for non-flood attacks.

\paragraph{Operational alerts and false positives.}
Baseline traffic was not alert-free: legitimate Snap7 sessions triggered Setup Communication alerts (\texttt{sid:1000001}). These operational false positives are predictable and manageable through procedural context---known engineering hosts, maintenance windows, correlation with change tickets---rather than through signature deletion alone. Removing Setup rules would blind the sensor to session-establishment floods and to the first phase of many reconnaissance scripts. The more sustainable approach is to treat Setup alerts as contextual signals and to prioritise follow-up when they co-occur with Write Var, Start/Stop, or elevated Ack\_Data rates from the same source within a short interval.

\paragraph{Block transfer and MITM behaviour.}
Block upload and download rules responded to the full handshake sequence when reference traffic was replayed through the sensor path, including direction-dependent Download Block legs where the PLC initiates \texttt{0x1b} toward the client (\texttt{sid:1000021}) and the client acknowledges with Ack\_Data at offset~19 (\texttt{sid:1000026}). MITM detection required combining write alerts (\texttt{sid:1000010}, \texttt{sid:1000011}) with a secondary indicator of elevated Ack\_Data volume (\texttt{sid:1000013}); no single opcode uniquely identifies response forgery because the interceptor preserves valid S7 headers while altering payload bytes. The MITM scenario therefore illustrates a broader lesson: some ICS attacks are visible only through correlation across multiple rules or network layers, not through one magic signature.

\paragraph{Malformed traffic and stack fidelity.}
Malformed-traffic rules fired on structural faults even when OpenPLC closed sessions early and did not expose every injected variant to the full parser chain. Partial coverage across the nine fault types still satisfied the detection criterion because at least one \texttt{100005x} rule matched in each campaign run. This behaviour reflects both the strength of encapsulation-level signatures (wrong TPKT version, length mismatch, missing COTP) and the limit of testing on a software PLC stack rather than on every Siemens firmware revision.

\paragraph{Sensor capacity.}
The IDS kept pace with mirrored traffic in the lab: cumulative decoded packets and alerts grew through the attack batch, but the kernel capture path reported no drops. This does not establish a maximum throughput bound for production mirror ports; it shows only that the chosen VM hardware, capture mode, and rule set were adequate for the thesis traffic profile. Capacity testing at line rate would be a separate exercise and is noted under future work.

\section{Limitations}

The results must be read within explicit boundaries. Ignoring these limits would overstate what a passive S7comm rule set can guarantee outside the laboratory.

\paragraph{Scale and duration of the testbed.}
The testbed scales to one PLC, one attacker, and short observation windows. It does not reproduce multi-site topologies, redundant paths, VLANs with partial mirroring, or weeks of continuous HMI polling. Conclusions about alert density and operator workload therefore describe the lab campaign---for example the short-window baseline density $\rho_{\mathrm{baseline}}$---not a running plant with three shifts of engineering activity and seasonal production changes.

\paragraph{Signature-based, intent-blind detection.}
Detection is signature-based and intent-blind. Write Var, Start/Stop, and block-transfer rules fire on authorized engineering actions as well as on attacks. A maintenance engineer loading a program through TIA Portal produces the same opcodes as an adversary replaying a captured Start PDU. Without asset inventories, allow lists, or correlation with change tickets, alerts require human triage. The thesis deliberately accepts this trade-off to preserve visibility into dangerous opcodes; context enrichment is left to operational processes or future correlation layers.

\paragraph{OpenPLC vs.\ Siemens hardware.}
OpenPLC implements S7comm through a software stack that differs from Siemens CPUs in error handling, session limits, and response to malformed input~\cite{openplc_s7_driver}. Malformed-traffic and DoS experiments reflect that stack: some PDUs were rejected at the TCP layer before Suricata saw a complete decode, and flood thresholds that worked in the lab may need revision against a physical S7-300/1200/1500 with different queue depths. Field devices may accept, reject, or crash on different inputs; the rule logic is portable, but alert rates are not.

\paragraph{Block transfer via PCAP replay.}
Block-transfer evaluation used recorded Siemens traffic replayed offline rather than live program loads through OpenPLC or TIA Portal. It validates rules against authentic PDUs and direction semantics but does not prove identical alert timing or session identifiers on operational engineering workflows. OpenPLC loads programs via REST, and S7-1500 environments often expose S7comm-plus, which Wireshark cannot decode with the same dissector used for classic S7comm~\cite{industrialmonitordirect_s7comm_plus,wireshark_s7comm}.

\paragraph{MITM rule semantics.}
MITM rule \texttt{1000013} measures Ack\_Data response rate from the PLC toward the HMI subnet, not modified payload content. Confirming forgery still depends on packet comparison, host forensics, ARP anomaly detection, or switch logs---techniques outside the S7comm rule set developed here. The rule supports triage (write plus abnormal read-response volume) but does not replace deep inspection of response data.

\paragraph{Visibility and encryption.}
Passive mirroring assumes the attacker crosses the monitored segment. Routed attacks that never traverse the SPAN port, compromised engineering stations on unmonitored VLANs, VPN tunnels, or encrypted overlays would evade the sensor regardless of signature quality. The thesis scope is a single mirrored Controller-layer segment; extending visibility is an architecture question, not a rule-syntax question.

\paragraph{Safety and response.}
The thesis does not model incident response, automatic blocking, or safety-system interaction. Suricata in this work is strictly an observer. Translating alerts into control actions---isolating a PLC, tripping a safety interlock, or blocking TCP/102 inline---raises separate safety and availability questions that depend on plant policy and often require human approval. Any IPS mode on OT links was intentionally out of scope.

\section{Directions for future work}

Several extensions would strengthen both the testbed and the detection approach. Each item below expands a limitation identified above into a concrete research or deployment path.

\paragraph{Longitudinal baseline and tuning.}
Running benign HMI and engineering traffic over days or weeks would yield realistic alert-density distributions and support data-driven threshold tuning for DoS and MITM-correlation rules. The 60-second baseline in Chapter~5 is sufficient to show that thresholds can sit above poll rate, but production SOC playbooks need seasonal and shift-aware baselines. Automated collection of Setup and Read SZL rates per engineering host would allow dynamic thresholds instead of fixed counts such as 80 Job PDUs per ten seconds.

\paragraph{Live block transfer on target hardware.}
Repeating upload and download evaluation on PLCSIM Advanced or physical S7 hardware would close the gap between PCAP replay and operational program-management workflows. The same rule set could be validated under live TIA Portal sessions, including S7comm-plus on newer controllers where PUT/GET compatibility modes affect visibility~\cite{industrialmonitordirect_s7comm_plus}. Success would be measured not only by SID hits but by alert ordering relative to engineering steps (Request Download before Download Block, and so on).

\paragraph{Context enrichment.}
Integrating asset lists, maintenance schedules, and multi-signal correlation would reduce operational false positives without weakening attack detection. Examples include suppressing Write Var alerts from known engineering IPs during approved windows, elevating priority when a new host sends Start/Stop, or joining S7 alerts with ARP spoofing indicators from the MITM scenario. A lightweight correlation engine above Suricata EVE JSON logs would be enough for a pilot without replacing the signature layer.

\paragraph{Coverage expansion.}
Additional scenarios---credential abuse on engineering stations, lateral movement from IT segments, combined IT/OT chains, and abuse of other Siemens services---would test whether the current SID grouping remains maintainable or requires hierarchical rule organisation. MITRE ATT\&CK for ICS~\cite{mitre_ics_attack} provides a taxonomy for mapping new rules to tactics and techniques so that coverage gaps can be reported systematically rather than ad hoc.

\paragraph{Sensor hardening and capacity.}
Fuzzing the IDS parser, measuring maximum sustainable mirror throughput, and evaluating Rust-based dissectors in Suricata would address resilience and performance limits before high-volume deployment. Kernel drop counters should be monitored continuously on mirror ports that carry full-duplex HMI polling plus concurrent engineering sessions. Stress tests with multi-gigabit capture and larger rule sets would establish headroom beyond the thesis VM configuration.

\paragraph{Operational integration.}
Mapping alerts to MITRE ATT\&CK for ICS, defining SOC playbooks per SID group, and piloting passive IDS on isolated plant segments would move the rule set from laboratory validation toward practical monitoring practice. Playbooks might specify, for example, that concurrent \texttt{1000010} and \texttt{1000013} from one attacker host triggers containment of that host while the PLC continues running until operations confirms process impact. Such integration requires OT--IT coordination and is the natural next step after technical feasibility is established.

In summary, the thesis demonstrates an end-to-end path from ICS attack simulation on S7comm to a tested Suricata rule set on a mirrored laboratory segment. The work confirms technical feasibility for the scenarios studied---reconnaissance through MITM---while leaving scale, context, hardware fidelity, and operational integration as the main avenues for continued research. A practitioner can reuse the rule structure and offset conventions directly; adapting thresholds, allow lists, and mirror placement to a specific site remains site-specific engineering work.

\end{document}
